DESING AND SIMULATION SOFTWARE FOR RHODOTRON TYPE ELECTRON ACCELERATORS

### 2023-08-17 msthesis.tex
changes from Erkcan & Gokhan hocas

% APPROVED BY PAGE
\supervisor{Prof. Veysi Erkcan Özcan}
\cosuperi{Gökhan Ünel, Ph.D.}
%\cosuperii{Title and Name of Cosupervisor II}
\examineri{Prof. Erhan Gülmez}
\examinerii{Bora Işıldak , Ph.D.}
%\examineriv{}
%\examinerv{}
\dateofapproval{DD.MM.YYYY}

@@

% ABS PAGE
\begin{abstract}
One page abstract will come here.  
\end{abstract}

% OZET PAGE
\begin{ozet}
Bir sayfa uzunluğunda özet gelecektir.
\end{ozet}

@@

@@

@@

% START OF THEORY

\chapter{THEORY}

@@

\subsection{Relation between momentum and acceleration}
In classical mechanics, Newton's second law defines force and states 
\begin{equation} \label{eq:newton_second_law}
    \vecthreeBF{F} = \frac{\partial \vecthreeBF{p}}{\partial t} = m \frac{\partial \vecthreeBF{v}}{\partial t} = m \vecthreeBF{a}
\end{equation}
In special relativity however, relativistic momentum is defined as $\vecthree{p} = \gamma m_0 \vecthree{v} $ where $\gamma$ is the Lorentz Factor;
\begin{equation*}
    \gamma = \frac{1}{\sqrt{1-v^2 / c^2}} = \frac{1}{\sqrt{1-\beta^2}}
\end{equation*}
Considering these two statements, we can find the relation between momentum and acceleration as follows
\begin{eqnarray*}
    \parDeriv{\vecthree{p}}{t} &=& m_0 \parDeriv{(\gamma \vecthree{v})}{t} = m_0 \{  \parDeriv{\gamma}{t}\vecthree{v} + \gamma \parDeriv{\vecthree{v}}{t}  \} \\
    \parDeriv{\gamma}{t} &=& \gamma^3 \dotProd{\beta}{\parDeriv{\beta}{t}} = \frac{\gamma^3}{c} \dotProd{\beta}{a} \\
    \parDeriv{\vecthree{p}}{t}  &=& m_0 \{   \frac{\gamma^3}{c} (\dotProd{\beta}{a} )\vecthree{v} + \gamma \parDeriv{\vecthree{v}}{t}  \} 
\end{eqnarray*}
\begin{equation} \label{eq:Frel}
    \vecthreeBF{F}  = \gamma m_0 \{ \vecthreeBF{a} + \gamma^2(\dotProd{\beta}{\textbf{a}})\vecthree{\beta} \}
\end{equation}
It is clear that acceleration is not necessarily parallel to the force anymore.

To start seperating the parallel and perpendicular components relative to $\vecthree{\beta}$, we can find $\vecthreeBF{a}_{||}$ and $\vecthreeBF{F}_{||}$.
\begin{equation} \label{eq:a_F_parallels}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{(\dotProd{a}{\beta})}{\beta^2}\beta 
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{||} = \frac{(\dotProd{F}{\beta})}{\beta^2}\beta 
    \end{aligned}
\end{equation}
\begin{eqnarray*}
    \dotProd{\textbf{F}}{\beta} &=& \gamma m_0 \{ \dotProd{\textbf{a}}{\beta} + \gamma^2(\dotProd{\beta}{\textbf{a}})\beta^2 \} \\
                                &=& \gamma m_0 (\dotProd{\textbf{a}}{\beta}) \{ \gamma^2\beta^2  + 1\} 
\end{eqnarray*}
Using $\gamma^2 \beta^2 + 1 = \gamma^2 $ we have,
\begin{equation*}
    \dotProd{\textbf{F}}{\beta} = m_0 \gamma^3 (\dotProd{\textbf{a}}{\beta} )
\end{equation*}
Inserting this
\begin{eqnarray}
    \vecthreeBF{F}_{||} &=& \frac{(\dotProd{\textbf{F}}{\beta})}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \frac{(\dotProd{\textbf{a}}{\beta} )}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \label{eq:Frel_parallel}
\end{eqnarray}
Therefore from \fromeqs{Frel}{Frel_parallel}
\begin{eqnarray*}
    \vecthreeBF{F}  &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \vecthreeBF{a} \\
                    &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \{ \vecthreeBF{a}_{||} +  \vecthreeBF{a}_{\perp} \} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma \{ \gamma^2\beta^2 + 1 \} + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma^3 + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& \vecthreeBF{F}_{||} + m_0 \gamma \vecthreeBF{a}_{\perp}
\end{eqnarray*}
We finally have two seperate equations which are in similar form with \fromeq{newton_second_law}.
\begin{equation} \label{eq:Frel_para_and_perp}
    \begin{aligned}
        \vecthreeBF{F}_{||} = \gamma^3  m_0 \vecthreeBF{a}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp}
    \end{aligned}
\end{equation}

@@

\subsection{Relativistic Lorentz Force}
Similar to non-relativistic version, relativistic Lorentz Force is given by the following 4-vector equality
\begin{equation}
    \frac{\partial p^{\mu}}{\partial \tau} = q F^{\mu \nu} u_{\nu}
\end{equation}
Where $ \partial \tau = \partial t / \gamma $,
\begin{equation}
    \begin{aligned}
        p^{\mu} = 
        \begin{bmatrix}
            W/c \\
            p_x         \\
            p_y \\
            p_z
        \end{bmatrix}    
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        F^{\mu\nu} = 
        \begin{bmatrix}
                0       & -E_x/c   & -E_y/c    & -E_z/c \\
                E_x/c   &   0      & -B_z      & B_y     \\
                E_y/c   &   B_z    &  0        & -B_x     \\
                E_z/c   &   -B_y   & B_x       & 0   
        \end{bmatrix} 
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        u_{\nu} = \gamma
        \begin{bmatrix}
                 c \\
                -v_x \\
                -v_y \\
                -v_z \\
        \end{bmatrix} 
    \end{aligned}
\end{equation}
Where $W$ is the energy of the particle and $\gamma$ is the lorentz factor mentioned in the previous section.

@@

\subsection{Acceleration caused by lorentz force}
Due to the nature of the cross product, lorentz force caused by a magnetic field is always perpendicular to the velocity of the particle. 
Therefore the acceleration of the magnetic field is straightforward
\begin{equation*}
    \vecthreeBF{F}_B = \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp} = \gamma  m_0\vecthreeBF{a}_{B}
\end{equation*}
The same thing cannot be said about electric field however. 
It can create force in any direction with respect to velocity. Therefore, we have the following equality;
\begin{equation} \label{eq:para_and_perp_acc_of_lorentz_force}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{q }{\gamma^3 m_0} \vecthreeBF{E}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{\perp} = \frac{q}{\gamma m_0} \{ \vecthreeBF{E}_{\perp} + \crossProdBF{v}{B}   \}
    \end{aligned}    
\end{equation}
Acceleration due to electric field can be simplified as;
\begin{eqnarray*}
    \vecthreeBF{a}_{\{B=0\}} = \vecthreeBF{a}_{E}   &=&  \vecthreeBF{a}_{||} + \vecthreeBF{a}_{\perp \{B=0\}} \\
                                &=& \frac{q}{m_0 \gamma} \{ \frac{\vecthreeBF{E}_{||}}{\gamma^2} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \{1 - \beta^2\}\vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp}-\beta^2\vecthreeBF{E}_{||} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E} - \beta^2\vecthreeBF{E}_{||} \} \\
\end{eqnarray*}
Using the fact that $\vecthreeBF{E}_{||} = \vec{\beta}(\dotProd{\textbf{E}}{\beta})/\beta^2 $, we finally have
\begin{equation} \label{eq:acc_of_E_and_B}
    \begin{aligned}
        \vecthreeBF{a}_{E} = \frac{q}{ \gamma m_0} \{ \vecthreeBF{E} - \vecthreeBF{v}\frac{(\dotProdBF{E}{v})}{c^2} \}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{B} = \frac{q}{\gamma m_0} ( \crossProdBF{v}{B} )
    \end{aligned}
\end{equation}

@@

At their core, particle accelerators utilize electromagnetic fields to impart energy to particles and control their trajectories. 
These fields are generated by intricate arrangements of magnets and RF (radiofrequency) cavities within the accelerator structure. 
By precisely controlling these fields, accelerators can propel particles to speeds close to the speed of light, enabling them to acquire high energies.

@@

The acceleration process in accelerators involves multiple stages. Initially, particles are injected into the accelerator at a relatively low energy. 
As they progress through the accelerator, they are subjected to alternating electric fields that accelerate them, while magnetic fields guide their trajectories.
Focusing elements, such as magnetic lenses or quadrupole magnets, ensure the particles remain tightly controlled.

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.

Particle accelerators are essential tools for scientific research. They enable scientists to probe the fundamental constituents of matter, study particle interactions, and explore the laws of physics. 
Accelerators have been instrumental in discovering and characterizing fundamental particles, such as quarks, leptons, and the Higgs boson. 
They also facilitate the production of high-intensity beams for applications in material science, radiotherapy, and industrial processes like particle irradiation and sterilization.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[scale=0.75]{./figures/to_be_added.png}
    \caption{A linear proton accelerator in KAHVELab}
\end{figure}
\subsection{Key Concepts}

@@

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
% TODO : ADD RF CAVITY PICTURE HERE
\begin{figure}[H]
    \centering
    \includegraphics[scale=0.75]{./figures/to_be_added.png}
    \caption{A rf cavity used in KAHVELab}
\end{figure}

@@

According to the Lorentz Force Law ( $\textit{Section \ref{sec:lorentz-force}}$ ), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.
% TODO : ADD STEERER MAGNET PICTURE HERE
\begin{figure}[H]
    \centering
    \includegraphics[scale=0.75]{./figures/to_be_added.png}
    \caption{A bending magnet used in KAHVELab}
\end{figure}

@@

Rhodotron Accelerator is a type of particle accelerator that was proposed by \textit{Jacques POTTIER} in his publication \textit{A new type of rf electron accelerator:The Rhodotron} in 1989 \cite{rhodo_pottier}. 
First prototype was built at CEA Saclay later in 1992 \cite{rhodo_prototype}. Its named after the greek word for rose, \textit{rhodos}, due to the shape of the design \cite{rhodos}.

The design of a rhodotron mainly consists of a coaxial cylindrical RF cavity and bending magnets surrounding it. RF cavity is fed by an external RF source, accelerating the electrons entering from an attached electron gun.

@@

Energy gain after each pass can be calculated by the following formula \cite{rhodo_pottier}
\begin{equation}
    \label{eq:W_gain_each_pass_pottier}
    W_1 = Z_{sp}^{1/2} P^{1/2} cos \phi  
\end{equation}
Where $P$ is the dissipated power, $\phi$ is the phase lag, taken as $15^\circ$ \cite{rhodo_pottier}.

Considering $Z_{sp} \propto \lambda^{1/2}$, \,\,\, $W_1 \propto \lambda^{1/4}$, \,\,\,   $V \propto \lambda^3$, where $V$ is the volume of the cavity, implementing the $p=1$ design in \fromfig{pottier_table1} is much more space efficient.
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/pottier_table2.png}
    \caption{Energy of a synchronous electron after each pass for both $p=1$ and $p=2$ \cite{rhodo_pottier}}
    \label{fig:pottier_table2}
\end{figure}
Total energy gain after n passes $W$ can then be found by \fromeq{W_gain_each_pass_pottier}, taking $\phi = 15^\circ$, $p=1$, i.e $R_2 = 0.27 \lambda$.
\begin{equation}
    \label{eq:W_total_gain_pottier}
    W \approx 2.14 \lambda^{1/4} P^{1/2} n \quad keV
\end{equation}

\begin{description}
    \item[First acceleration] Electrons in the rhodotron cavity are accelerated by the electric field created between two coaxial cylinders, towards the inner cylinder when they are ejected into the cavity. See \fromfig{rhod_cycle_1}
    \item[Inner shielding] While inside the inner cylinder, the cylinder acts as a faraday cage and shields the electrons inside while the electric field is being reversed.See \fromfig{rhod_cycle_2}
    \item[Second acceleration] Once the electrons leave the inner cylinder, they accelerate towards the outer cylinder by the reversed electric field until they leave the cavity. See \fromfig{rhod_cycle_3}
    \item[Magnets] After leaving the cavity, an electromagnet placed in their path steers the electrons back into the cavity in which time the electric field changes the direction again. See \fromfig{rhod_cycle_4}
\end{description}

>>>>

% APPROVED BY PAGE
\supervisor{Prof. Veysi Erkcan Özcan}
\cosuperi{Assoc. Prof. Gökhan Ünel}
%\cosuperii{Title and Name of Cosupervisor II}
\examineri{Asst. Prof. Bora Akgün}
\examinerii{Assoc. Prof. Bora Işıldak}
%\examineriv{}
%\examinerv{}
\dateofapproval{DD.MM.YYYY}

@@

% ABS PAGE
\begin{abstract}
    The focus of this study is to design and optimize rose-pattle-type electron accelerators; a new simulation tool called \textit{Rhodotron Simulation} has been developed for this purpose.
    A coaxial cavity with an operating frequency of $107.5$ MHz is currently being manufactured in the Kandilli Detector Accelerator and Instrumentation Laboratory (KAHVELab).
    This new tool's capabilities will be tested with the cavity and will be used to design complementary bending magnets for reaching $1-5$ MeV beam energy.
\end{abstract}

% OZET PAGE 
\begin{ozet}
Bu çalışma, gül yaprağı tipi elektron hızlandırıcıları tasarlamak ve optimize etmek üzerine odaklanmaktadır. 
Bu amaçla \textit{Rhodotron Simulation} adında yeni bir benzetim uygulaması geliştirilmiş, $107.5$ MHz hedef frekansında çalışmak üzere tasarlanmış bir koaksiyal kavitesinin üretimi, Kandilli Algıç Hızlandırıcı ve Enstrümentasyon Laboratuvarı'nda devam etmektedir.
Yeni benzetim uygulaması, kabiliyetleri üretilen kavite ile test edildikten sonra bükme mıknatısları tasarım ve optimizasyonunda $1-5$ MeV demet enerjisine ulaşmak için kullanılacaktır.
\end{ozet}

@@

% LIST OF SYMBOLS PAGE
\begin{symbols}
    % The title will be typeset as "LIST OF SYMBOLS".
    % Use a separate \sym command for each symbols definition.
    % First, Latin symbols in alphabetical order
    \sym{$\vec{a}$}{Acceleration}
    \sym{\textbf{$\vec{B}$}}{Magnetic field}
    \sym{c}{Speed of light in vacuum}
    \sym{E}{Energy}
    \sym{\textbf{$\vec{E}$}}{Electric field}
    \sym{f}{Frequency}
    \sym{$\vec{F}$}{Force}
    \sym{$\vec{p}$}{Momentum}
    \sym{$\vec{r}$}{Position}
    \sym{t}{Time}
    \sym{T}{Period}
    \sym{$\vec{v}$}{Velocity}
    \sym{Z}{Shunt Impedance}
    % 1 EMPTY LINE BETWEEN LATIN AND GREEK SYMBOLS GROUP!!!
    \sym{}{}
    % Then Greek symbols in alphabetical order
    \sym{$\gamma$}{Lorentz Factor}
    \sym{$\beta$}{Beta Factor}
    \sym{$\phi_{lag}$}{Phase Lag}

    \sym{$\mu X$}{Mean Value of X}
    \sym{$\sigma X$}{Standard Deviation of X}
    \sym{}{}
\end{symbols}

% LIST OF ABBREV PAGE
\begin{abbreviations}
    % Abbreviations in alphabetical order
   \sym{2D}{One Dimensional}
   \sym{2D}{Two Dimensional}
   \sym{3D}{Three Dimensional}
   \sym{eV}{Electron Volt}
   \sym{Hz}{Hertz}
   \sym{KAHVELab}{Kandilli Detector Accelerator and \\Instrumentation Laboratory}
   \sym{Linac}{Linear Particle Accelerator}
   \sym{ODE}{Ordinary Differential Equation}
   \sym{RF}{Radio Frequency}
   \sym{s}{Second}
   \sym{T}{Tesla}
   \sym{V}{Volt}

\end{abbreviations}

@@

\newcommand{\fromch}[1]{\textit{chapter \ref{ch:#1}}}

@@

% START OF INTRODUCTION AND THEORY

\chapter{INTRODUCTION AND THEORY}

Particle accelerators are essential tools for scientific research. They enable scientists to probe the fundamental constituents of matter, study particle interactions, and explore the laws of physics. 
Accelerators have been instrumental in discovering and characterizing fundamental particles, such as quarks, leptons, and more recently the Higgs boson. 
They also facilitate the production of high-intensity particle beams for applications in industrial processes. A conservative estimate puts the number of industrial high-current electron
accelerators to be in the thousands, worldwide.

MeV-level electron beams, both as is and as X-Ray sources, are especially useful for industrial and medical applications. 
For example, high-energy beams (5–10 MeV) are used in medical accelerators and provide services such as medical device sterilization and food irradiation;
with medium energy devices (400 keV–5 MeV) wires, cables and tires are improved, factory flue gases or wastewater can be treated.
Although most of the industrial electron accelerators operating at MeV-level are linear accelerators at the moment, due to their compact nature, circular accelerators have been gaining prominence.

Rose-pattle accelerator is a type of circular accelerator suitable for producing electron beams at MeV-level energy. 
It that was developed in 1989 and was named \textit{Rhodotron} \cite{rhodo_pottier}. This name was trademarked by \textit{IBA Group} in the following years.

@@

\subsection{Relation between momentum and acceleration}
In classical mechanics, Newton's second law defines force, $\vecthreeBF{F}$, and states 
\begin{equation} \label{eq:newton_second_law}
    \vecthreeBF{F} = \frac{\partial \vecthreeBF{p}}{\partial t} = m \frac{\partial \vecthreeBF{v}}{\partial t} = m \vecthreeBF{a}
\end{equation}
where $\vecthreeBF{p}$ is the momentum, $m\vecthreeBF{v}$ of the particle. In special relativity however, relativistic momentum is defined as $\vecthree{p} = \gamma m_0 \vecthree{v} $ where $\gamma$ is the Lorentz Factor:
\begin{equation}
    \gamma = \frac{1}{\sqrt{1-v^2 / c^2}} = = \frac{1}{\sqrt{1-\beta^2}}
            = \frac{1}{\sqrt{1- (\dotProd{v}{v})/ c^2}} =  \frac{1}{\sqrt{1- \dotProd{\beta}{\beta}}}
\end{equation}
Considering these two statements, we can find the relation between momentum and acceleration as follows:
\begin{eqnarray}
    \parDeriv{\vecthree{p}}{t} &=& m_0 \parDeriv{(\gamma \vecthree{v})}{t} = m_0 \{  \parDeriv{\gamma}{t}\vecthree{v} + \gamma \parDeriv{\vecthree{v}}{t}  \} \\
    \parDeriv{\gamma}{t} &=& \gamma^3 \dotProd{\beta}{\parDeriv{\beta}{t}} = \frac{\gamma^3}{c} \dotProd{\beta}{a} \\
    \parDeriv{\vecthree{p}}{t}  &=& m_0 \{   \frac{\gamma^3}{c} (\dotProd{\beta}{a} )\vecthree{v} + \gamma \parDeriv{\vecthree{v}}{t}  \} 
\end{eqnarray}
thus,
\begin{equation} \label{eq:Frel}
    \vecthreeBF{F}  = \gamma m_0 \{ \vecthreeBF{a} + \gamma^2(\dotProd{\beta}{\textbf{a}})\vecthree{\beta} \}
\end{equation}
It is clear that acceleration is not necessarily parallel to the force anymore.

To start separating the parallel and perpendicular components relative to $\vecthree{\beta}$, we can find $\vecthreeBF{a}_{||}$ and $\vecthreeBF{F}_{||}$.
\begin{equation} \label{eq:a_F_parallels}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{(\dotProd{a}{\beta})}{\beta^2}\beta 
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{||} = \frac{(\dotProd{F}{\beta})}{\beta^2}\beta 
    \end{aligned}
\end{equation}
\begin{eqnarray}
    \dotProd{\textbf{F}}{\beta} &=& \gamma m_0 \{ \dotProd{\textbf{a}}{\beta} + \gamma^2(\dotProd{\beta}{\textbf{a}})\beta^2 \} \\
                                &=& \gamma m_0 (\dotProd{\textbf{a}}{\beta}) \{ \gamma^2\beta^2  + 1\} 
\end{eqnarray}
Using $\gamma^2 \beta^2 + 1 = \gamma^2 $ we have,
\begin{equation}
    \dotProd{\textbf{F}}{\beta} = m_0 \gamma^3 (\dotProd{\textbf{a}}{\beta} )
\end{equation}
inserting this into \fromeq{a_F_parallels},
\begin{eqnarray}
    \vecthreeBF{F}_{||} &=& \frac{(\dotProd{\textbf{F}}{\beta})}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \frac{(\dotProd{\textbf{a}}{\beta} )}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \label{eq:Frel_parallel}
\end{eqnarray}
Therefore from \fromeqs{Frel}{Frel_parallel}
\begin{eqnarray}
    \vecthreeBF{F}  &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \vecthreeBF{a} \\
                    &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \{ \vecthreeBF{a}_{||} +  \vecthreeBF{a}_{\perp} \} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma \{ \gamma^2\beta^2 + 1 \} + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma^3 + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& \vecthreeBF{F}_{||} + m_0 \gamma \vecthreeBF{a}_{\perp}
\end{eqnarray}
We finally have two separate equations which are in similar form with \fromeq{newton_second_law}.
\begin{equation} \label{eq:Frel_para_and_perp}
    \begin{aligned}
        \vecthreeBF{F}_{||} = \gamma^3  m_0 \vecthreeBF{a}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp}
    \end{aligned}
\end{equation}

@@

\subsection{Relativistic Lorentz Force}
Similar to non-relativistic version, relativistic Lorentz Force is given by the following 4-vector equality
\begin{equation}
    \frac{\partial p^{\mu}}{\partial \tau} = q F^{\mu \nu} u_{\nu}
\end{equation}
where $ \partial \tau = \partial t / \gamma $ and greek letters are Lorentz indices,
\begin{equation}
    \begin{aligned}
        p^{\mu} = 
        \begin{bmatrix}
            W/c \\
            p_x \\
            p_y \\
            p_z
        \end{bmatrix}    
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        F^{\mu\nu} = 
        \begin{bmatrix}
                0       & -E_x/c   & -E_y/c    & -E_z/c \\
                E_x/c   &   0      & -B_z      & B_y    \\
                E_y/c   &   B_z    &  0        & -B_x   \\
                E_z/c   &   -B_y   & B_x       & 0   
        \end{bmatrix} 
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        u_{\nu} = \gamma
        \begin{bmatrix}
                 c \\
                -v_x \\
                -v_y \\
                -v_z \\
        \end{bmatrix} 
    \end{aligned}
\end{equation}
where $W$ is the energy of the particle and $\gamma$ is the Lorentz factor mentioned in the previous section.

@@

\subsection{Acceleration caused by Lorentz force}
Due to the nature of the cross product, Lorentz force caused by a magnetic field is always perpendicular to the velocity of the particle. 
Therefore the acceleration of the magnetic field is straightforward
\begin{equation*}
    \vecthreeBF{F}_B = \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp} = \gamma  m_0\vecthreeBF{a}_{B}
\end{equation*}
The same thing cannot be said about electric field however. 
It can create force in any direction with respect to velocity. Therefore, we have the following equality;
\begin{equation} \label{eq:para_and_perp_acc_of_lorentz_force}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{q }{\gamma^3 m_0} \vecthreeBF{E}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{\perp} = \frac{q}{\gamma m_0} \{ \vecthreeBF{E}_{\perp} + \crossProdBF{v}{B}   \}
    \end{aligned}    
\end{equation}
Acceleration due to electric field can be simplified as;
\begin{eqnarray*}
    \vecthreeBF{a}_{\{B=0\}} = \vecthreeBF{a}_{E}   &=&  \vecthreeBF{a}_{||} + \vecthreeBF{a}_{\perp \{B=0\}} \\
                                &=& \frac{q}{m_0 \gamma} \{ \frac{\vecthreeBF{E}_{||}}{\gamma^2} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \{1 - \beta^2\}\vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp}-\beta^2\vecthreeBF{E}_{||} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E} - \beta^2\vecthreeBF{E}_{||} \} \\
\end{eqnarray*}
Using the fact that $\vecthreeBF{E}_{||} = \vec{\beta}(\dotProd{\textbf{E}}{\beta})/\beta^2 $, we finally have
\begin{equation} \label{eq:acc_of_E_and_B}
    \begin{aligned}
        \vecthreeBF{a}_{E} = \frac{q}{ \gamma m_0} \{ \vecthreeBF{E} - \vecthreeBF{v}\frac{(\dotProdBF{E}{v})}{c^2} \}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{B} = \frac{q}{\gamma m_0} ( \crossProdBF{v}{B} )
    \end{aligned}
\end{equation}
where $\vecthreeBF{a}_{E}$ consists of two components in the direction of $\vecthreeBF{E}$ and $\vecthreeBF{v}$.
The component in the direction of $\vecthreeBF{v}$ depends on the $\dotProdBF{E}{v}$, 
resulting in $\vecthreeBF{a}_{E} || \vecthreeBF{E}$ if $\vecthreeBF{E} \perp \vecthreeBF{v}$.

@@

At their core, particle accelerators utilize electromagnetic fields to impart energy to particles and control their trajectories. 
These fields are generated by intricate arrangements of electromagnets and RF (radiofrequency) cavities within the accelerator structure. 
By precisely controlling these fields, accelerators can propel particles to speeds close to the speed of light, thus to high energies.

@@

The acceleration process in accelerators involves multiple stages. Initially, particles are injected into the accelerator at a relatively low energy. 
As they progress through the accelerator, they are subjected to electric fields that accelerate them, while magnetic fields guide their trajectories.
Focusing elements, such as solenoid, dipole or quadrupole magnets, ensure the particles remain tightly controlled.

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/kahvelab_linac.png}
    \caption{A linear proton accelerator in KAHVELab \cite{kahvelab_linac}}
\end{figure}
\subsection{Key Concepts}

@@

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
% TODO : ADD RF CAVITY PICTURE HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.75\textwidth]{./figures/pill_box.jpeg}
    \caption{An RF cavity used in KAHVELab}
\end{figure}

@@

According to the Lorentz Force Law ( $\textit{Section \ref{sec:lorentz-force}}$ ), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.

@@

Rhodotron Accelerator is a type of particle accelerator that was proposed by \textit{Jacques POTTIER} in 1989 \cite{rhodo_pottier}. 
First prototype was built at CEA Saclay later in 1992 \cite{rhodo_prototype}. It is named after the greek word for rose, \textit{rhodos}, due to the shape of the design \cite{rhodos}.

The design of a rhodotron mainly consists of a coaxial cylindrical RF cavity and bending magnets surrounding it. RF cavity is fed by an external RF source, accelerating the electrons entering from an attached electron injector.

\begin{description}
    \item[First acceleration] Electrons in the rhodotron cavity are accelerated by the electric field created between two coaxial cylinders, towards the inner cylinder when they are ejected into the cavity. (\fromfig{rhod_cycle_1})
    \item[Inner shielding] While inside the inner cylinder, the cylinder acts as a faraday cage and shields the electrons inside while the electric field is being reversed. (\fromfig{rhod_cycle_2})
    \item[Second acceleration] Once the electrons leave the inner cylinder, they accelerate towards the outer cylinder by the reversed electric field until they leave the cavity. (\fromfig{rhod_cycle_3})
    \item[Recirculating Magnets] After leaving the cavity, an electromagnet placed in their path steers the electrons back into the cavity in which time the electric field changes the direction again. (\fromfig{rhod_cycle_4})
\end{description}

This cycle can be repeated as long as real world constraints such as; placements and dimensions of the electromagnets, power requirements due to increasing magnetic field in order for sharper turns, can be overcomed.
After the desired amount of cycles, also called passes, has been completed, the electrons exit the accelerator.

This process is explained further in the \fromfigf{rhod_cycle_1}{rhod_cycle_2}{rhod_cycle_3}{rhod_cycle_4} where $T$ is the period of the electric field.

@@

Energy gain after each pass can be calculated by the following formula \cite{rhodo_pottier}
\begin{equation}
    \label{eq:W_gain_each_pass_pottier}
    W_1 = Z_{sp}^{1/2} P^{1/2} \cos \phi  
\end{equation}
where $P$ is the dissipated power, $\phi$ is the phase lag, typically taken as $15^\circ$ \cite{rhodo_pottier}.

Considering $Z_{sp} \propto \lambda^{1/2}$, \,\,\, $W_1 \propto \lambda^{1/4}$, \,\,\,   $V \propto \lambda^3$, where $V$ is the volume of the cavity, implementing the $p=1$ design in \fromfig{pottier_table1} is much more space efficient.
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/pottier_table2.png}
    \caption{Energy of a synchronous electron after each pass for both $p=1$ and $p=2$ \cite{rhodo_pottier}}
    \label{fig:pottier_table2}
\end{figure}
Total energy gain after $n$ passes, $W$, can then be found by \fromeq{W_gain_each_pass_pottier}, taking $\phi = 15^\circ$, $p=1$, i.e $R_2 = 0.27 \lambda$.
\begin{equation}
    \label{eq:W_total_gain_pottier}
    W \approx 2.14 \lambda^{1/4} P^{1/2} n \quad keV
\end{equation}
